---
title: "Data Modeling and Imputation for Composite AERA Study"
author: 
- name: "Nirmal Ghimire, Ph.D."
  url: https://www.linkedin.com/in/nirmal-ghimire-5b96a034/
  affiliation: K-16 Literacy Center at University of Texas at Tyler
  affiliation_url: https://www.uttyler.edu/education/literacy-center/
  orcid_id: 0000-0002-2032-1624
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_float: true
    code_folding: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      comment = NA,
                      warning = FALSE,
                      message = FALSE,
                      tidy = 'styler',
                      error = FALSE, 
                      highlight = TRUE, 
                     prompt = FALSE,
                     cache = TRUE)
```

```{r library}
library(haven)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(janitor)
#library(car)# For Levene's Test
library(compute.es)# For effect sizes
library(effects) # for adjusted means
library(pastecs) # for descriptive statistics
#library(multcomp) # For post hoc test
#library(WRS2) # for robust tests
#library(lme4) # To conduct Hierarchical Modeling
#library(lmerTest) # To provide p-values in type I, II, or III anova summary tables for linear mixed models
library(geomtextpath)
library(lavaan)
library(lavaanPlot)
library(tidySEM)
library(semTable)
library(mice)
library(ggmice)
library(broom.mixed)
```

## A. Loading Data and Getting them Ready for Analysis

### 1. School, Student Background, and Student Home ICT Data Together

```{r student_school_data}
# Loading Student and School Data
ss_data <- read.csv("C:/Users/nghimire/OneDrive - University of Texas at Tyler/Documents/edsurvey_PISA_USA/SS_final_data.csv", 
                      as.is = TRUE)
# Changing character vectors to factors
ss_data <- ss_data%>%
  mutate_if(is.character, as.factor)|>
  dplyr::select(GENDER,ETHNICITY,LANGN,COB_MOM,COB_DAD,MISCED,FISCED,COM_HOM,INTERNET,ICTHOME,SCH_LOCA,SCH_TYPE,FRPL,SCH_FRPL,SCH_ELS,SCH_DISA,EXT_CLAS,READ_SCR)

# Getting rid of unnecessary variables


#names(ss_data)
str(ss_data)
summary(ss_data)
```

### Checking for Assumptions of Multivariate Normality and Outliers for SEM Study

#### Mardia's Test of Multivariate Normality

```{r mardia_ss}
ss_num <- dplyr::select(ss_data, SCH_ELS, SCH_FRPL, ICTHOME, READ_SCR, SCH_DISA)
library(QuantPsyc)
mult.norm(ss_num)$mult.test

ss_num$ss_maha <- mahalanobis(ss_num, colMeans(ss_num), cov(ss_num))
ss_num$maha_p <- pchisq(ss_num$ss_maha, df=17, lower.tail=FALSE)
ss_high_maha <- ss_num%>%
  filter(maha_p <= .001)%>%
  arrange(maha_p)

ss_high_maha

nrow(ss_high_maha)
```

### Checking missing data pattern

```{r imputing_ss_data}
VIM::aggr(ss_data, col=c('dodgerblue','maroon'), numbers=TRUE, sortVars=TRUE, labels=names(ss_data), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))
#summary(checking_missing_data, plot=FALSE)

naniar::mcar_test(ss_data)

# Imputing
new_ssdata <- mice::mice(ss_data[], m = 5, maxit = 50, method = "pmm", seed = 0411, printFlag = FALSE)
```

### Plotting the imputed vs original data

```{r checking_the_imputed_ssdata}
densityplot(new_ssdata, ~COB_MOM)
ggmice::ggmice(
  new_ssdata, 
  ggplot2::aes(x = SCH_FRPL, y = READ_SCR))+
ggplot2::geom_jitter(width = 0.25)+
  ggplot2::labs( x = "School FRPL Percentage")
```

### Pooled {lm} on New SS Data

```{r testing_new_data}
# Running lm model on Pooled data
ss_pooled <- with(new_ssdata, lm(READ_SCR ~ SCH_ELS + FRPL + EXT_CLAS + SCH_DISA + SCH_FRPL + SCH_TYPE + SCH_LOCA + INTERNET + ICTHOME + FISCED + COB_DAD + COB_MOM + MISCED + COM_HOM + ETHNICITY + LANGN + GENDER))
summary(pool(ss_pooled))
```

### Breaking Imputed Data in Individual Sets

```{r individual_data_set}
ss_data_1 <- complete(new_ssdata, 1)
ss_data_2 <- complete(new_ssdata, 2)
ss_data_3 <- complete(new_ssdata, 3)
ss_data_4 <- complete(new_ssdata, 4)
ss_data_5 <- complete(new_ssdata, 5)
```

### LM on Dataset 1

```{r lm_ss1}
ss_1 <- with(ss_data_1, lm(READ_SCR ~ SCH_ELS + FRPL + EXT_CLAS + SCH_DISA + SCH_FRPL + SCH_TYPE + SCH_LOCA + INTERNET + ICTHOME + FISCED + COB_DAD + COB_MOM + MISCED + COM_HOM + ETHNICITY + LANGN + GENDER))
#summary(ss_1)
```

### LM on Dataset 2

```{r lm_ss2}
ss_2 <- with(ss_data_2, lm(READ_SCR ~ SCH_ELS + FRPL + EXT_CLAS + SCH_DISA + SCH_FRPL + SCH_TYPE + SCH_LOCA + INTERNET + ICTHOME + FISCED + COB_DAD + COB_MOM + MISCED + COM_HOM + ETHNICITY + LANGN + GENDER))
#summary(ss_2)
```

### LM on Dataset 3

```{r lm_ss3}
ss_3 <- with(ss_data_3, lm(READ_SCR ~ SCH_ELS + FRPL + EXT_CLAS + SCH_DISA + SCH_FRPL + SCH_TYPE + SCH_LOCA + INTERNET + ICTHOME + FISCED + COB_DAD + COB_MOM + MISCED + COM_HOM + ETHNICITY + LANGN + GENDER))
#summary(ss_3)
```

### LM on Dataset 4

```{r lm_ss4}
ss_4 <- with(ss_data_4, lm(READ_SCR ~ SCH_ELS + FRPL + EXT_CLAS + SCH_DISA + SCH_FRPL + SCH_TYPE + SCH_LOCA + INTERNET + ICTHOME + FISCED + COB_DAD + COB_MOM + MISCED + COM_HOM + ETHNICITY + LANGN + GENDER))
#summary(ss_4)
```

### LM on Dataset 5

```{r lm_ss5}
ss_5 <- with(ss_data_5, lm(READ_SCR ~ SCH_ELS + FRPL + EXT_CLAS + SCH_DISA + SCH_FRPL + SCH_TYPE + SCH_LOCA + INTERNET + ICTHOME + FISCED + COB_DAD + COB_MOM + MISCED + COM_HOM + ETHNICITY + LANGN + GENDER))
#summary(ss_5)
summary(ss_data_5)
str(ss_data_5)
write.csv(ss_data_5, file = "s_data.csv", row.names = FALSE)
#save(ss_data_5, file = "s_data.RData")
```

### 2. Teacher Data

```{r teacher_data}
teacher_final <- read.csv("C:/Users/nghimire/OneDrive - University of Texas at Tyler/Documents/edsurvey_PISA_USA/t_final.csv", 
                      as.is = TRUE)
    #names(teacher_final)
# Changing class of character variables
teacher_final <- data.frame(lapply(teacher_final, function(b) if(is.character(b)) as.factor(b) else b))

# Getting rid of unnecessary column
teacher_final <- teacher_final[,-c(1,8,12,13)]

#summary(teacher_final)
#str(teacher_final)

# Teacher Education Data
teacher_education <- read.csv("C:/Users/nghimire/OneDrive - University of Texas at Tyler/Documents/edsurvey_PISA_USA/teacher_education.csv", 
                      as.is = TRUE)
options(scipen = 999)
# Getting rid of unnecessary column
teach_edu <- teacher_education[,-1]
# Changing the class of teacher_education 
#teach_edu$teacher_education <- as.factor(teach_edu$teacher_education)


# Getting Rid of First Four Digits from 'teacher_id' and 'school_id'
teach_edu$school_id <- stringr::str_sub(teach_edu$CNTSCHID, -4,-1)
teach_edu$teacher_id <- stringr::str_sub(teach_edu$CNTTCHID, -4,-1)

# Getting Rid of Leading Zeros
teach_edu$school_id <- stringr::str_remove(teach_edu$school_id, "^0+")
teach_edu$teacher_id <- stringr::str_remove(teach_edu$teacher_id, "^0+")

# Changing the class
teach_edu <- teach_edu|>
  dplyr::select(school_id, teacher_id, teacher_education)|>
  mutate(school_id = as.integer(school_id),
         teacher_id = as.integer(teacher_id),
         highest_education = as.factor(teacher_education))

#str(teach_edu)

# Merging the data sets
t_data <- merge(teach_edu, teacher_final, by = c("school_id","teacher_id"))
t_data <- t_data|>
  dplyr::select(school_id, teacher_id, gender, age, teacher_type, full_time, highest_education, teaching_exp, teacher_edu, initial_qual, prof_dev, reading_score)

str(t_data)
summary(t_data)
```

### Mardia Test for Teacher Data

```{r mardia_teacher}
teach_numeric <- dplyr::select(t_data, age, teaching_exp, reading_score)
mult.norm(teach_numeric)$mult.test
## Mahalanobis Distance
teach_numeric$mahal <- mahalanobis(teach_numeric, colMeans(teach_numeric), cov(teach_numeric))
teach_numeric$p <- pchisq(teach_numeric$mahal, df=3, lower.tail=FALSE)
high_maha_teacher <- teach_numeric%>%
  dplyr::filter(p <= .001)%>%
  arrange(p)
high_maha_teacher
#rowsum(high_maha_teacher)
```

## Imputing the data in teacher_education column

```{r imputing_data}
options(scipen = 999)
# Plotting the missing pattern
    #plot_pattern(t_data)
# Or
checking_missing_data <- VIM::aggr(t_data, col=c('dodgerblue','maroon'), numbers=TRUE, sortVars=TRUE, labels=names(t_data), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))
summary(checking_missing_data, plot=FALSE)

naniar::mcar_test(t_data)

# Imputing
new_data <- mice::mice(t_data, m = 5, maxit = 50, method = "pmm", seed = 0410, printFlag = FALSE)
#summary(new_data)
```

### Plotting the imputed vs original data

```{r checking_the_imputed_data}
densityplot(new_data, ~highest_education)
ggmice::ggmice(
  new_data, 
  ggplot2::aes(x = teaching_exp, y = reading_score))+
ggplot2::geom_jitter(width = 0.25)+
  ggplot2::labs( x = "Teaching Experience")
```

### Pooled Linear Model

```{r running_lm}
pooled_lm <- with(new_data, lm(reading_score ~ gender+age+teacher_type+full_time+highest_education+teaching_exp+teacher_edu+initial_qual+prof_dev))
## Pooled Statistics
summary(pool(pooled_lm))
```

### Saving all the imputed data set as individual dataset

```{r breaking_individual_data}
teacher_data_1 <- complete(new_data, 1)
teacher_data_2 <- complete(new_data, 2)
teacher_data_3 <- complete(new_data, 3)
teacher_data_4 <- complete(new_data, 4)
teacher_data_5 <- complete(new_data, 5)
```

### Using Imputed 1

```{r lm_1}
lm_one <- with(teacher_data_1, lm(reading_score ~ gender+age+teacher_type+full_time+highest_education+teaching_exp+teacher_edu+initial_qual+prof_dev))
#summary(lm_one)
```

### Using Imputed 2

```{r lm_2}
lm_two <- with(teacher_data_2, lm(reading_score ~ gender+age+teacher_type+full_time+highest_education+teaching_exp+teacher_edu+initial_qual+prof_dev))
summary(lm_two)
#names(teacher_data_2)
str(teacher_data_2)
#summary(teacher_data_2)
#The decision is to use the second set of imputed data to run further analysis
#write.csv(teacher_data_2, file = "t_data.csv", row.names = FALSE)
```

### Using Imputed 3

```{r lm_3}
lm_three <- with(teacher_data_3, lm(reading_score ~ gender+age+teacher_type+full_time+highest_education+teaching_exp+teacher_edu+initial_qual+prof_dev))
#summary(lm_three)
```

### Using Imputed 4

```{r lm_4}
lm_four <- with(teacher_data_4, lm(reading_score ~ gender+age+teacher_type+full_time+highest_education+teaching_exp+teacher_edu+initial_qual+prof_dev))
#summary(lm_four)
```

### Using Imputed 5

```{r lm_5}
lm_five <- with(teacher_data_5, lm(reading_score ~ gender+age+teacher_type+full_time+highest_education+teaching_exp+teacher_edu+initial_qual+prof_dev))
#summary(lm_five)
```

## Average Reading Scores Based on Student and Teacher Characteristics: Categorical Variables Having More than Two Groups Only

### Teacher Characteristics

### Teacher Highest Education

```{r, include=FALSE}
teacher_data <- teacher_data_2|>
  mutate(high_education = recode(highest_education,
                                    "Associate’s degree" = 'High School; Some College; Associates Degree',
                                    "High school and/or some college courses" = 'High School; Some College; Associates Degree'))
### Getting Rid of highest_education variable
  teacher_data$highest_education <- NULL

### Calculating average scores
teacher_data|>
  dplyr::group_by(high_education)|>
  summarise(mean_score = mean(reading_score),
            sd_score = sd(reading_score))
```

### Teacher Education

```{r, include=FALSE}
teacher_data|>
  dplyr::group_by(teacher_edu)|>
  summarise(mean_score = mean(reading_score),
            sd_score = sd(reading_score))
```

### Initial Qualification

```{r}
teacher_data <- teacher_data|>
   mutate(init_qual = recode(initial_qual,
                             "Other" = 'Training on Other Pedagogical Profession or Other',
                             "Training on Another Pedagogical Profession" = 'Training on Other Pedagogical Profession or Other'))
### Getting Rid of initial_qual variable
  teacher_data$initial_qual <- NULL

# Reading Score
  teacher_data|>
  dplyr::group_by(init_qual)|>
  summarise(mean_score = mean(reading_score),
            sd_score = sd(reading_score))
```

### Regression for Teacher

```{r reg_teach}
t_model <- lm(reading_score ~ high_education + init_qual + gender + age + teacher_type + full_time + teaching_exp + teacher_edu + prof_dev, data = teacher_data)
summary(t_model)
#QuantPsyc::lm.beta(t_model)
```

### Student/School Characteristics

```{r}
s_data <- ss_data_5|>
  dplyr::mutate(student_race = recode(ETHNICITY,
                          "Other" = "Two or More Race/Other",
                          "Two or More Race" = "Two or More Race/Other"))
### Getting Rid of ETHNICITY
s_data$ETHNICITY <- NULL

## Reading Scores
s_data|>
  dplyr::group_by(student_race)|>
  summarise(mean_score = mean(READ_SCR),
            sd_score = sd(READ_SCR))
```

### Mother Education(MISCED)

```{r}
s_data <- s_data|>
  dplyr::mutate(mom_education = recode(MISCED,
                          "ISCED 1" = "ISCED 1/None",
                          "None" = "ISCED 1/None"))
### Getting Rid of MISCED
s_data$MISCED <- NULL

## Reading Scores
s_data|>
  dplyr::group_by(mom_education)|>
  summarise(mean_score = mean(READ_SCR),
            sd_score = sd(READ_SCR))
```

### Father Education(FISCED)

```{r}
s_data <- s_data|>
  dplyr::mutate(dad_education = recode(FISCED,
                          "ISCED 1" = "ISCED 1/None",
                          "None" = "ISCED 1/None"))
### Getting Rid of MISCED
s_data$FISCED <- NULL

## Reading Scores
s_data|>
  dplyr::group_by(dad_education)|>
  summarise(mean_score = mean(READ_SCR),
            sd_score = sd(READ_SCR))
```

### School Location (SCH_LOCA)

```{r}
s_data <- s_data|>
  dplyr::mutate(school_location = recode(SCH_LOCA,
                          "rural" = "small_town/rural",
                          "small_town" = "small_town/rural"))
### Getting Rid of MISCED
s_data$SCH_LOCA <- NULL

## Reading Scores
s_data|>
  dplyr::group_by(school_location)|>
  summarise(mean_score = mean(READ_SCR),
            sd_score = sd(READ_SCR))
```

### Home Langauge

```{r}
## Reading Scores
s_data|>
  dplyr::group_by(LANGN)|>
  summarise(mean_score = mean(READ_SCR),
            sd_score = sd(READ_SCR))
```

### Home Internet

```{r}
## Reading Scores
s_data|>
  dplyr::group_by(INTERNET)|>
  summarise(mean_score = mean(READ_SCR),
            sd_score = sd(READ_SCR))
```

### Free or Reduced Price Lunch Status

```{r}
## Reading Scores
s_data|>
  dplyr::group_by(FRPL)|>
  summarise(mean_score = mean(READ_SCR),
            sd_score = sd(READ_SCR))
summary(s_data)
```

### Regression SS
#### A. Combined Model
```{r}
model_ss <- lm(READ_SCR ~ GENDER + student_race + LANGN + COB_MOM + COB_DAD + mom_education + dad_education + COM_HOM + INTERNET + ICTHOME + school_location + SCH_TYPE + FRPL + SCH_ELS + SCH_DISA + EXT_CLAS, data = s_data)
summary(model_ss)
#QuantPsyc::lm.beta(model_ss)
```


### B. School Model
```{r school_model_lm}
options(scipen = 999)
model_school <- lm(READ_SCR ~ school_location + SCH_TYPE + FRPL + SCH_ELS + SCH_DISA + EXT_CLAS, data = s_data)
summary(model_school)
```


### C. Student and ICT Model
```{r student_ICT_Model}
model_stu_ICT <- lm(READ_SCR ~ GENDER + student_race + LANGN + COB_MOM + COB_DAD + mom_education + dad_education + COM_HOM + INTERNET + ICTHOME, data = s_data)
summary(model_stu_ICT)
```

### D. ICT Model
```{r ICT_only_model}
ict_only_model <- lm(READ_SCR ~ COM_HOM + INTERNET + ICTHOME, data = s_data)
summary(ict_only_model)
```

### D. Student Only Model
```{r student_only_model}
student_only_model <- lm(READ_SCR ~ GENDER + student_race + LANGN + COB_MOM + COB_DAD + mom_education + dad_education, data = s_data)
summary(student_only_model)
```
