---
title: Teachers' Impact on Student Reading Scores Using Teacher Survey, Part II
author: 
- name: "Nirmal Ghimire, Ph.D."
  url: https://www.linkedin.com/in/nirmal-ghimire-5b96a034/
  affiliation: K-16 Literacy Center at University of Texas at Tyler
  affiliation_url: https://www.uttyler.edu/education/literacy-center/
  orcid_id: 0000-0002-2032-1624
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      comment = NA,
                      warning = FALSE,
                      message = FALSE,
                      tidy = 'styler',
                      error = FALSE, 
                      highlight = TRUE, 
                     prompt = FALSE)
library(haven)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(car)# For Levene's Test
library(compute.es)# For effect sizes
library(effects) # for adjusted means
library(pastecs) # for descriptive statistics
library(multcomp) # For post hoc test
library(WRS2) # for robust tests
library(lme4) # To conduct Hierarchical Modeling
library(lmerTest) # To provide p-values in type I, II, or III anova summary tables for linear mixed models
library(geomtextpath)
```


```{r reading_score_teacher_employment}
# Loading Files
teacher_final <- read.csv("C:/Users/nghimire/OneDrive - University of Texas at Tyler/Documents/edsurvey_PISA_USA/teacher_data_final.csv", 
                      as.is = TRUE)
    str(teacher_final)
    #summary(teacher_final)
    
teacher_imputed <- read.csv("C:/Users/nghimire/OneDrive - University of Texas at Tyler/Documents/edsurvey_PISA_USA/imputed_teacher_data.csv", 
                      as.is = TRUE)
    str(teacher_imputed)
```

```{r changing_labels}
# Teacher Data
teacher_final$teacher_id <- as.factor(teacher_final$teacher_id)
teacher_final$gender <- factor(teacher_final$gender, levels = c(0,1), labels = c("Female", "Male"))
teacher_final$teacher_type <- factor(teacher_final$teacher_type, levels = c(0,1), labels = c("Reading-Language Teacher", "General Teacher"))
teacher_final$full_time <- factor(teacher_final$full_time, levels = c(0,1), labels = c("Full-Time", "Part-Time"))
teacher_final$emp_stat <- factor(teacher_final$emp_stat, levels = c(0,1,2,3), labels = c("Full-Time", "Part-Time, 71-90% of Full-Time Hours", "Part-Time, 50-70% of Full-Time Hours", "Part-Time, less than 50% of Full-Time Hours"))
teacher_final$teacher_edu <- factor(teacher_final$teacher_edu, levels = c(0, 1, 2), labels = c("None. No Teacher Education", "Program of 1 Year or Less", "Yes, Program Longer than 1 Year"))
teacher_final$initial_qual <- factor(teacher_final$initial_qual, levels = c(0,1,2,3,4), labels = c("Pre-service: Strandard Teacher Training Program", "In-service: Teacher Training Program", "Work-Based Teacher Training", "Training on Another Pedagogical Profession", "Other"))
teacher_final$training <- factor(teacher_final$training, levels = c(0,1), labels = c("Participated", "Didn't Participate"))
teacher_final$workshop <- factor(teacher_final$workshop, levels = c(0,1), labels = c("Participated", "Didn't Participate"))
teacher_final$prof_dev <- factor(teacher_final$prof_dev, levels = c(0,1), labels = c("Participated", "Didn't Participate"))

summary(teacher_final)
#write.csv(teacher_final, "t_final.csv")
  
# Imputed Data
teacher_imputed$teacher_id <- as.factor(teacher_imputed$teacher_id)
teacher_imputed$gender <- factor(teacher_imputed$gender, levels = c(0,1), labels = c("Female", "Male"))
teacher_imputed$teacher_type <- factor(teacher_imputed$teacher_type, levels = c(0,1), labels = c("Reading-Language Teacher", "General Teacher"))
teacher_imputed$full_time <- factor(teacher_imputed$full_time, levels = c(0,1), labels = c("Full-Time", "Part-Time"))
teacher_imputed$emp_stat <- factor(teacher_imputed$emp_stat, levels = c(0,1,2,3), labels = c("Full-Time", "Part-Time, 71-90% of Full-Time Hours", "Part-Time, 50-70% of Full-Time Hours", "Part-Time, less than 50% of Full-Time Hours"))
teacher_imputed$teacher_edu <- factor(teacher_imputed$teacher_edu, levels = c(0, 1, 2), labels = c("None. No Teacher Education", "Program of 1 Year or Less", "Yes, Program Longer than 1 Year"))
teacher_imputed$initial_qual <- factor(teacher_imputed$initial_qual, levels = c(0,1,2,3,4), labels = c("Pre-service: Strandard Teacher Training Program", "In-service: Teacher Training Program", "Work-Based Teacher Training", "Training on Another Pedagogical Profession", "Other"))
teacher_imputed$training <- factor(teacher_imputed$training, levels = c(0,1), labels = c("Participated", "Didn't Participate"))
teacher_imputed$workshop <- factor(teacher_imputed$workshop, levels = c(0,1), labels = c("Participated", "Didn't Participate"))
teacher_imputed$prof_dev <- factor(teacher_imputed$prof_dev, levels = c(0,1), labels = c("Participated", "Didn't Participate"))

summary(teacher_imputed)
```

## Various Teachers' Characteristics and Students' Reading Score
### a. Teacher Education
```{r means_by_teacher_education_types}
teacher_final|>
  na.omit()|>
    group_by(teacher_edu)|>
  summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score),
            total_teacher = n())
```


### b. Teachers' Gender
```{r gender}
teacher_final|>
  na.omit()|>
    group_by(gender)|>
  summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score),
            total_teacher = n())
```

### c. Teacher Type
```{r teacher_type}
teacher_imputed|>
  na.omit()|>
    group_by(teacher_type)|>
  summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score),
            total_teacher = n())
```

### d. Teaching Status
```{r teaching_status}
teacher_final|>
  na.omit()|>
    group_by(full_time)|>
  summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score),
            total_teacher = n())
```

### e. Teaching Status Imputed Data
```{r imputed_teaching_status}
teacher_imputed|>
      group_by(full_time)|>
  summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score),
            total_teacher = n())
```

### f. Employment Status
```{r employment_status}
teacher_final|>
  na.omit()|>
    group_by(emp_stat)|>
  summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score),
            total_teacher = n())
```

### g. Initial Qualification
```{r initial_qualification}
teacher_final|>
  na.omit()|>
    group_by(initial_qual)|>
  summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score),
            total_teacher = n())
```

### g. Teacher Training Opportunities in Last 12-Months' Period
```{r teacher_training}
teacher_final|>
  na.omit()|>
    group_by(training)|>
  summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score),
            total_teacher = n())
```

### h. Workshop Opportunities in Last 12-Months' Period
```{r workshop}
teacher_final|>
  na.omit()|>
    group_by(workshop)|>
  summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score),
            total_teacher = n())
```

### i. Professional Development Opportunity in Last 12-Months' Period
```{r professional_development}
teacher_final|>
  na.omit()|>
    group_by(prof_dev)|>
  summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score),
            total_teacher = n())
```

### Teacher Education and Students' Reading Scores
```{r visualization_1}
teacher_edu_density <- teacher_imputed|>
  ggplot(aes(x = reorder(teacher_edu, reading_score), y = reading_score))+
    geom_boxplot(fill = "white", colour = "blue", varwidth = TRUE)+
  geom_hline(yintercept = 505, label = "National Average", hjust = 1, vjust = 0.5, size = .3)+
  labs(title = "Comparative Reading Scores Based on Teachers' Education",
    x = "Teacher Education",
       y = "Reading Scores")+
  geom_text(label = "National Average",
             x = 1.4, y = 370,
            family = "serif",
            size = 2.3)+ 
  geom_segment(aes(x = 1.4, y = 400, xend = 1.4, yend = 500), 
               linewidth = 0.06,
               arrow = arrow(length = unit(.05, 'cm')))+
  theme(legend.position = "none")+
  coord_flip()+
      theme_classic()
teacher_edu_density  
#ggsave("teacher_education_vs_reading_scores.tiff", units="in", width=6.5, height=4, #dpi=300, compression = 'lzw')
```

### Teacher Type, Teaching Experience, and Students' Readign Scores
```{r teacher_type_&_reading_scores}
library(dendextend)
english_teacher <- subset(teacher_imputed, teacher_type == "Reading-Language Teacher")
#head(english_teacher)
general_teacher <- subset(teacher_imputed, teacher_type == "General Teacher")
#head(general_teacher)

et_box <- ggplot(teacher_imputed, 
                 aes(teacher_type, reading_score))+
  geom_boxplot(alpha = 0.5, width = 0.3) + 
  stat_summary(fun = mean, size = 2, geom = "point") +
  stat_summary(fun = mean, geom = "text", vjust = -0.6,
               aes(label = paste(round(..y.., digits = 2))))+
  scale_x_discrete(labels = c("Reading Teacher", "General Teacher"))+
      ggpubr::theme_classic2()
      #et_box


library(geomtextpath)
te_scat <- ggplot(teacher_imputed, 
       aes(x = reading_score, y = teaching_exp, color = teacher_type, label = teacher_type))+
    geom_labelsmooth(method = "lm", boxlinewidth = 0) + 
  labs(title = "",
    x = "Reading Scores",
       y = "Teaching Experience")+
  theme_classic() + 
  theme(legend.position = "none")
# Putting Plots Together
comined_plot <- ggpubr::ggarrange(et_box, te_scat,
                  ncol = 2, align = "h",
                  common.legend = FALSE)
comined_plot
#ggsave("teacher_type_experience_reading_scores.tiff", units="in", width=6.5, height=4, #dpi=300, compression = 'lzw')
```

### Hierarchical Linear Models
#### Null Models
```{r linear_model}
hlm_null <- nlme::gls(reading_score ~ 1, data = teacher_final, method = "ML")

hlm_null_intercept <- nlme::lme(reading_score ~ 1, data = teacher_imputed, random = ~1|school_id, method = "ML")

anova(hlm_null, hlm_null_intercept)
```

#### Null Model with Imputed Data 
```{r null_model_1}
hlm_null_1 <- nlme::gls(reading_score ~ 1, data = teacher_imputed, method = "ML")

summary(hlm_null_1)
```

#### Null Model with Random Intercept
```{r null_model_with_random_intercept}
hlm_null_1_intercept <- nlme::lme(reading_score ~ 1, data = teacher_imputed, random = ~1|school_id, method = "ML")

summary(hlm_null_1_intercept)
```

#### Model Comparison
```{r checking_for_hlm_fit}
anova(hlm_null_1,hlm_null_1_intercept)
```


### Teacher Null
```{r teacher_null}
teacher_null <- nlme::gls(reading_score ~ 1, data = teacher_imputed, method = "ML")
summary(teacher_null)

teacher_null_rintercept <- nlme::lme(reading_score ~ 1, data = teacher_imputed, random = ~1|teacher_id, method = "ML")
summary(teacher_null_rintercept)

anova(teacher_null,teacher_null_rintercept)
```


### HLM Full
```
{r hlm_gender}
hlm_full <- lmer(reading_score ~ gender + age + teacher_type + full_time + teaching_exp + teacher_edu + initial_qual + training + workshop + prof_dev + (gender + age + teacher_type + full_time + teaching_exp + teacher_edu + initial_qual + training + workshop + prof_dev| school_id), data = teacher_imputed)

summary(hlm_full)

#effectsize::standardize_parameters(hlm_full, method = "pseudo")

#confint(hlm_full)
```

### Quantile Regression for Visualiztion
```{r quantile_regression}
library(quantreg)
quant_1 <- rq(formula = reading_score ~ teaching_exp, tau=c(.05, .25, .50, .75, .95), data = teacher_imputed, method = "fn")
summary(quant_1)
# Plotting the Output and Saving as .pdf
pdf("teaching_experience_readin_score.pdf", width = 6.5, height = 7)
plot(quant_1)
dev.off()
```

*"nid" for SE are computed using the Hall-Sheather bandwidth rule. We can use "ker" instead, if we want Powell kernel version, "boot" for bootstrap, etc. (visit: https://cran.r-project.org/web/packages/quantreg/vignettes/rq.pdf)*


```{r quantile_two}
teacher_imputed$reading_score <- as.numeric(as.factor(teacher_imputed$reading_score))
teacher_imputed$teaching_exp <- as.numeric(as.factor(teacher_imputed$teaching_exp))
attach(teacher_imputed)
png("quantile_experience_vs_reading_scores.png")
plot(reading_score, teaching_exp, xlab = "Students' Reading Scores", ylab = "Teacher Experience in Years", type = "n", xaxt = "n")
tau_s <- c(.1, .25, .75, .95)

abline(rq(log10(reading_score)~log10(teaching_exp), tau = .5), col = "blue")
abline(lm(log10(reading_score)~log10(teaching_exp), lty = 3), col = "red")
for(i in 1:length(tau_s)){
  abline(rq(log10(reading_score)~log10(teaching_exp), tau = tau_s[i]), col = "gray")
}

dev.off()
```


### Teachers' Age and Average School Reading Scores
```{r quantile_three}
quant_3 <- rq(formula = reading_score ~ age, tau=c(0.05, .25, .50, .75, 0.95), data = teacher_final, method = "fn")
summary(quant_3)

plot(quant_3)

# Plotting the Output and Saving as .pdf
pdf("age_reading_score.pdf", width = 6.5, height = 7)
plot(quant_3)
dev.off()

#ggplot(teacher_final, aes(reading_score,age))+
    #geom_quantile(quantiles = c(0.05, .25, .50, .75, 0.95))+ #theme_classic()

plot(summary(quant_3), parm = "age")
```

### Age and Teacher Experience
```{r age_experience}
quant_4 <- rq(formula = age ~ teaching_exp, tau = c(.1, .25, 0.5, .75, .95), data = teacher_imputed)
summary(quant_4)
# Plotting the Output and Saving as .pdf
plot(quant_4)
```

### Reading Scores Description
```{r reading_score_only}
# (3526) reading scores
teacher_final|>
     summarise(M_EAN = mean(reading_score),
            S_D = sd(reading_score), 
            total_teacher = n())
# (1675 teachers)
teacher_final|>
  na.omit()|>
     summarise(M_EAN = mean(age),
            S_D = sd(age), 
            total_teacher = n())
# (1675 teachers)
teacher_final|>
  na.omit()|>
     summarise(M_EAN = mean(teaching_exp),
            S_D = sd(teaching_exp), 
            total_teacher = n())
```
